Hydra Battles and Cie

### 2021-08-09 doc/chapter-powers.tex
fix some small non-typo issues in chapter-powers.tex

Nothing looks simpler than writing some function for computing $x^n$.
On the contrary, this simple programming exercise allows us to address
advanced programming techniques such as:
\begin{itemize}
\item monadic programming, and continuation passing style
\item type classes, and generalized rewriting
\item proof engineering, in particular proof re-using
\item proof by reflection
\item polymorphism and parametricity
\item composition of correct programs, etc.
\end{itemize}

@@

In~\cite{BC04}, several exercises~\footnote{Exercises 9.8 (page 270), 9.10 (page 271), 9.15 (page 276), 9.17 (page 284), and 15.8 (page 418).}
present ways to compute Fibonacci numbers, with the less number of recursive calls  as possible. Please note that these optimizations and the formal proof of their correctness are \emph{ad-hoc}, \emph{i.e.} exclusively written for the 
Fibonacci numbers.
In contrast, the optimizations we present in this document apply, in their vast majority, \emph{generic} techniques of efficient computation of powers in a monoid. 
This example of Fibonacci numbers has been developed with Yves Bertot, who wrote a first version with \texttt{SSreflect/Mathcomp}~\cite{MCB}.

@@

\subsection{A common notation for multiplication}
\label{op-classes}
\index{coq}{Type classes!Operational type classes}
≈Ò
\emph{Operational type classes}~\cite{BS2011}
allow us to define a common notation 
for multiplication in any algebraic structure. 
First, we associate a class to the notion of \emph{multiplication} 
on any type $A$.

@@

In Sect.\vref{naive-matrix}, we defined a function for computing powers of any $2\times 2$ 
matrix over any semi-ring. For proving a simple property of matrix exponentiation, we had
 to prove that matrix multiplication is associative and admits the identity matrix as a neutral element. These properties are easily expressed within the type class framework, by defining a \emph{family} of monoids.
It suffices to define an instance of \texttt{Monoid} within the scope of an hypothesis
of type \texttt{semi\_ring\_theory}

@@

\section{Addition chains}
\index{maths}{Addition chains}
An \emph{addition chain} (In short : \emph{a chain})~\cite{brauer1939} is a representation of a sequence of
intermediate steps that lead to the evaluation of  $x^n$, under the 
assumption that each of these steps is a computation of  a power $x^i$, with 
$i<n$.

In articles from the combinatorist  community, 
\emph{e.g.}~\cite{brauer1939,DBLP:journals/ipl/BerstelB87},  addition chains 
are represented as sequences of positive integers, each member of which 
is either $1$ or  the sum of two previous elements.
For instance, the three following sequences are addition chains for the exponent $87$:

@@

It is possible to associate to any addition chain a directed acyclic graph:
whenever $i=j+k$, there is an arc from $x^j$ to $x^i$ and an arc
from $x^k$ to $x^i$. Figures~\ref{fig:chain-87-eucl}  and 
\ref{fig:chain-87-bin} show the graphical representations of 
 $c_{87}$  and $c'_{87}$. 
Please note that some chains may be represented by various different dags.
For instance, we can associate four different dags to the chain $(1,2,3,4,6,9,13)$.

@@

Thus, the main steps of a correctness proof of a given chain, \emph{e.g.}
\texttt{C87} will be the following ones:
\begin{enumerate}
\item generate a subgoal as in page~\pageref{fig:big-goal},
\item express each term of the equivalence as the image of a term
     of type \texttt{Monoid\_Exp $A$},
\item normalize both terms and verify that their normal forms are equal.
\end{enumerate}

>>>>

Nothing looks simpler than writing a function for computing $x^n$.
But on the contrary, this simple programming exercise allows us to address
advanced programming techniques such as:
\begin{itemize}
\item monadic programming, and continuation passing style
\item type classes, and generalized rewriting
\item proof engineering, in particular proof reuse
\item proof by reflection
\item polymorphism and parametricity
\item composition of correct programs, etc.
\end{itemize}

@@

In~\cite{BC04}, several exercises~\footnote{Exercises 9.8 (page 270), 9.10 (page 271), 9.15 (page 276), 9.17 (page 284), and 15.8 (page 418).}
present ways to compute Fibonacci numbers, with the less number of recursive calls  as possible. Please note that these optimizations and the formal proof of their correctness are \emph{ad-hoc}, \emph{i.e.}, exclusively written for the
Fibonacci numbers.
In contrast, the optimizations we present in this document apply, in their vast majority, \emph{generic} techniques of efficient computation of powers in a monoid. 
This example of Fibonacci numbers has been developed with Yves Bertot, who wrote a first version with \texttt{SSreflect/Mathcomp}~\cite{MCB}.

@@

\subsection{A common notation for multiplication}
\label{op-classes}
\index{coq}{Type classes!Operational type classes}

\emph{Operational type classes}~\cite{BS2011}
allow us to define a common notation 
for multiplication in any algebraic structure. 
First, we associate a class to the notion of \emph{multiplication} 
on any type $A$.

@@

In Sect.\vref{naive-matrix}, we defined a function for computing powers of any $2\times 2$ 
matrix over any semi-ring. For proving a simple property of matrix exponentiation, we had
 to prove that matrix multiplication is associative and admits the identity matrix as a neutral element. These properties are easily expressed within the type class framework, by defining a \emph{family} of monoids.
It suffices to define an instance of \texttt{Monoid} within the scope of a hypothesis
of type \texttt{semi\_ring\_theory}.

@@

\section{Addition chains}
\index{maths}{Addition chains}
An \emph{addition chain} (in short, a \emph{chain})~\cite{brauer1939} is a representation of a sequence of
intermediate steps that lead to the evaluation of  $x^n$, under the 
assumption that each of these steps is a computation of  a power $x^i$, with 
$i<n$.

In articles from the combinatorist  community, 
\emph{e.g.},~\cite{brauer1939,DBLP:journals/ipl/BerstelB87},  addition chains
are represented as sequences of positive integers, each member of which 
is either $1$ or  the sum of two previous elements.
For instance, the three following sequences are addition chains for the exponent $87$:

@@

It is possible to associate to any addition chain a directed acyclic graph:
whenever $i=j+k$, there is an arc from $x^j$ to $x^i$ and an arc
from $x^k$ to $x^i$. Figures~\ref{fig:chain-87-eucl}  and 
\ref{fig:chain-87-bin} show the graphical representations of 
 $c_{87}$  and $c'_{87}$. 
Please note that some chains may be represented by various different dags (directed acyclic graphs).
For instance, we can associate four different dags to the chain $(1,2,3,4,6,9,13)$.

@@

Thus, the main steps of a correctness proof of a given chain, \emph{e.g.},
\texttt{C87} will be the following ones:
\begin{enumerate}
\item generate a subgoal as in page~\pageref{fig:big-goal},
\item express each term of the equivalence as the image of a term
     of type \texttt{Monoid\_Exp $A$},
\item normalize both terms and verify that their normal forms are equal.
\end{enumerate}
